Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-08-11 demystify.tex
updates

\begin{enumerate}
  \item Should you apply to a PhD CS program in the US?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than in other countries (\autoref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty members (\autoref{sec:evalapps}).
          \item Individually faculty member \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so do contact and introduce yourself (\autoref{sec:contact}).
         \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are letters of recommendation (LORs),  statements of purpose (SOP), and research experience and potential, e.g., publications.
          \item \emph{LORs are very important}, but only if they are personalized and talk about your research ability (\autoref{sec:lor}).
          \item \emph{SOP is very important}. Write it in such a way that makes you \emph{stand out} (\autoref{sec:research-statement} and \autoref{sec:improve-your-chance})
          \item  \emph{Publications can greatly help}. Papers in good venues are concrete evidence that you have successfully engaged in research. But they are not required! (\autoref{sec:research-experiences}).

@@

\paragraph{Help your LOR Writer} As mentioned above, do not write your own letter and generic letters do not give much value. One thing you can do to help your writer to write a strong, customized LoR for you is to provide details or unique things about yourself. For example, let them know about your GPA, research and work experiences, papers (if any), or anything you want them to mention. You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.
Sometimes your writer will explicitly ask you for such information, but if not, you should provide it anyway (especially if you have not interacted with them much or have not done much research with them).

\paragraph{Remind Your Writers} After you submit your application, you should tell your writers about that and let them know they will soon receive an email from the university to submit their letters.  You should also remind them when you submit your application and ask them to submit their letters on time if they haven't done so.  Note that most places only have deadlines for the applicant, but are very flexible with the letter writers (in many cases do not even give them any deadline).  Also, many places do not begin the admission review process right after the deadline and work on application reviews in the next semester (mid-January).

@@

\paragraph{VEF 2.0} For Vietnamese students, it's worth mentioning the \href{https://vef2.org/}{VEF2.0 program}, which has helped many students gain admission to top CS PhD programs in the US. VEF2.0 invites US faculty members from leading institutions to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily just the interview), these reference letters are generally effective and have helped many students get into good PhD programs.

@@

This is a simple task often overlooked by many applicants: \textbf{tailor your SOP} to the institution you're applying for,
e.g., why do you apply \emph{here}? Provide names of professors who you're interested in (if they are not already in the adcom, your application might get forwarded to them for evaluation; and they might be interested in interviewing and recruiting you).
This shows that you're serious and have done homework on places you're applying to.
Adcom will look for this part (\autoref{sec:why-rejected}).

@@

\section{Research Experiences}\label{sec:research-experiences}

@@

\paragraph{Publications} The most concrete evidence of research ability is having \emph{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the you has been successfully involved in research. Publications are never required for application, however given the competitiveness of the process, they can significantly strengthen your application and are becoming the norm, e.g., those admitted to top schools often have multiple first-authored papers at top places.

@@

\paragraph{Work Experiences} Work experiences at \emph{well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you or your LoR writers should explicitly say something about them in your statement.

@@

\paragraph{Competitions} Winning \emph{internationally recognized competitions} can also demonstrate your research potential.
For example, participating in Math Olympiads if you want to do theory or winning ACM programming contests if you want to ``build'' systems, e.g., software analysis.

@@

\paragraph{GREs Do Not Matter} While a few schools still require taking the GRE (e.g., UCF), most good CS PhD programs in the US \textbf{no longer} require it. Thus, if you haven't taken the GRE or have bad scores, then don't waste time (re)taking it. It does not help much!

@@

\paragraph{English Tests (IELTS, TOEFL)} Unless your degrees are from specific countries such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-Mason#standard-tests-waiver-eligible-countries}{these}, you will need to take standardized English tests.

@@

\paragraph{Work Experiences} Work experiences, especially if they are not research-based, \emph{do not} carry much weight. Similarly, LoR from your supervisor for non-research experience will not count much.

Do not spend much time talking about non-research-related job in your SOP.
However, if you have research-based work experiences, then they can be useful and should be mentioned in your SOP and LORs. For example, if you have worked at a research lab, e.g., Microsoft, DoE, VinAI in Vietnam, etc (\autoref{sec:research-experiences}), then you should mention that in your SOP and ask your supervisor to write about your research experiences and potential in their LOR (\autoref{sec:lor}).

\paragraph{Online Courses and Certificates}
In general, these do not carry much weight as they do not show research ability. We do not care much if you have taken an online Coursera AI course or have a professional certificate from Microsoft.
However, as mentioned in \autoref{sec:non-stem}, if you do not have a CS background, you might be able to use these to show you have sufficient CS knowledge.

@@

You still can apply to PhD in CS \emph{as long as} you can demonstrate you are ready for it through your background, research experiences, LoRs, statements, etc. You might be even able to leverage this to make your profile stand out as mentioned in \autoref{sec:improve-your-chance}.

@@

\paragraph{Research Experiences \autoref{sec:research-opportunities}} You might have many opportunities to do research as an undergraduate, e.g., through REUs and internships at your undergrad university.  Highlight such experiences in your application.

@@

As discussed in \autoref{sec:research-opportunities}, research experience is a crucial component of a strong PhD application. It gives you opportunities to try out research, determine what research area you're interested in, publish papers (\autoref{sec:research-experiences}), connect with researchers, and get strong LoRs (\autoref{sec:lor}).

>>>>

\begin{enumerate}
  \item Should you apply to a PhD CS program in the US?
        \begin{itemize}
          \item \emph{Yes, definitely}.  CS PhD study in the US is fully funded and admission into good universities is not any harder than in other countries (\autoref{sec:should}).
        \end{itemize}
  \item How is your application evaluated?
        \begin{itemize}
          \item Applications are evaluated by the \emph{PhD Admission} (adcom) committee and each application is typically reviewed by three faculty members (\autoref{sec:evalapps}).
          \item Individually faculty member \emph{cannot directly admit} a student---so do not email and ask if you have a chance. However, faculty can \emph{advocate} for a student and therefore increase their admission chance---so do contact and introduce yourself (\autoref{sec:contact}).
         \end{itemize}
  \item Application Materials
        \begin{itemize}
          \item The committee will look at various factors, but the most important ones are letters of recommendation (LORs),  statements of purpose (SOP), and research experience and potential, e.g., publications.
          \item \emph{LORs are very important}, but only if they are personalized and talk about your research ability (\autoref{sec:lor}).
          \item \emph{SOP is very important}. Write it in such a way that makes you \emph{stand out} (\autoref{sec:research-statement} and \autoref{sec:improve-your-chance})
          \item  \emph{Publications can greatly help}. Papers in good venues are concrete evidence that you have successfully engaged in research. But they are not required! (\autoref{sec:research-experience}).

@@

\subsection{Help your LOR Writer} As mentioned above, do not write your own letter and generic letters do not give much value. One thing you can do to help your writer to write a strong, customized LoR for you is to provide details or unique things about yourself. For example, let them know about your GPA, research and work experience, papers (if any), or anything you want them to mention. You can also provide them with a draft of your SOP so that they can see what you are saying about yourself and complement that with their own perspective.
Sometimes your writer will explicitly ask you for such information, but if not, you should provide it anyway (especially if you have not interacted with them much or have not done much research with them).

\subsection{Remind Your Writers} After you submit your application, you should tell your writers about that and let them know they will soon receive an email from the university to submit their letters.  You should also remind them when you submit your application and ask them to submit their letters on time if they haven't done so.  Note that most places only have deadlines for the applicant, but are very flexible with the letter writers (in many cases do not even give them any deadline).  Also, many places do not begin the admission review process right after the deadline and work on application reviews in the next semester (mid-January).

@@

\subsection{VEF 2.0} For Vietnamese students, it's worth mentioning the \href{https://vef2.org/}{VEF2.0 program}, which has helped many students gain admission to top CS PhD programs in the US. VEF2.0 invites US faculty members from leading institutions to conduct rigorous interviews with VEF students and subsequently provide reference letters on their behalf. Despite the limited interaction between the interviewer and the interviewee (primarily just the interview), these reference letters are generally effective and have helped many students get into good PhD programs.

@@

This is a simple task often overlooked by many applicants: \textbf{tailor your SOP} to the institution you're applying for,
e.g., why do you apply \emph{here}? who do you want to work with?
Provide names of professors who you're interested in (if they are not already in the adcom, your application might get forwarded to them for evaluation; and they might be interested in interviewing and recruiting you).
This shows that you're serious and have done homework on places you're applying to.
Adcom will look for this part (\autoref{sec:why-rejected}).

@@

\section{Research Experience}\label{sec:research-experience}

@@

\subsection{Publications} The most concrete evidence of research ability is having \emph{papers in reputable international journals or conferences}.
Having published papers, especially at top venues, is a sign that the you has been successfully involved in research. Publications are never required for application, however given the competitiveness of the process, they can significantly strengthen your application and are becoming the norm, e.g., those admitted to top schools often have multiple first-authored papers at top places.

@@

\subsection{Work Experience} Work experience at \emph{well-known research laboratories}, such as Microsoft Research, can strengthen your
application.  Unfortunately, many good research places in your countries, e.g., VinAI in Vietnam, remain relatively unknown to most universities in the US. So you or your LoR writers should explicitly say something about them in your statement.

@@

\subsection{Competitions} Winning \emph{internationally recognized competitions} can demonstrate your research potential.
For example, participating in Math Olympiads if you want to do theory or winning ACM programming contests if you want to ``build'' systems, e.g., software analysis.

@@

\subsection{GREs Do Not Matter Much} While a few schools still require taking the GRE (e.g., UCF), most good CS PhD programs in the US \textbf{no longer} require it. Thus, if you haven't taken the GRE or have bad scores, then don't waste time (re)taking it. It does not help much!

@@

\subsection{English Tests (IELTS, TOEFL)} Unless your degrees are from specific countries such as \href{https://github.com/dynaroars/dynaroars.github.io/wiki/About-Mason#standard-tests-waiver-eligible-countries}{these}, you will need to take standardized English tests.

@@

\subsection{Work Experience} Work experience, especially if they are \emph{not} research-based, \emph{do not} carry much weight for PhD (they can help for MS application though). Similarly, LoR from your supervisor for non-research experience will not count much.

Do not spend much time talking about non-research-related job in your SOP.
However, if you have research-based work experiences, then they can be useful and should be mentioned in your SOP and LORs. For example, if you have worked at a research lab, e.g., Microsoft, DoE, VinAI in Vietnam, etc (\autoref{sec:research-experience}), then you should mention that in your SOP and ask your supervisor to write about your research experiences and potential in their LOR (\autoref{sec:lor}).

\subsection{Online Courses and Certificates}
In general, these do not carry much weight as they do not show research ability. We do not care much if you have taken an online Coursera AI course or have a professional certificate from Microsoft.
However, as mentioned in \autoref{sec:non-stem}, if you do not have a CS background, you might be able to use these to show you have sufficient CS knowledge.

@@

You still can apply to PhD in CS \emph{as long as} you can demonstrate you are ready for it through your background, research experience, LoRs, statements, etc. You might be even able to leverage this to make your profile stand out as mentioned in \autoref{sec:improve-your-chance}.

@@

\paragraph{Research Experience \autoref{sec:research-opportunities}} You might have many opportunities to do research as an undergraduate, e.g., through REUs and internships at your undergrad university.  Highlight such experience in your application.

@@

As discussed in \autoref{sec:research-opportunities}, research experience is a crucial component of a strong PhD application. It gives you opportunities to try out research, determine what research area you're interested in, publish papers (\autoref{sec:research-experience}), connect with researchers, and get strong LoRs (\autoref{sec:lor}).
